Speeding up decimal multiplication

### 2020-11-14 paper/paper.tex
Update

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{cite}

@@

@@

The following notation is employed throughout this section:
$$\langle x_0, \ldots, x_{n-1} \rangle [i] = x_i.$$

@@

Let us now fix a base $B \in \mathbb{N}, \, B > 1,$ and introduce the following notation:
$$\myllangle x_0, \ldots, x_{n-1} \myrrangle = \sum\limits_{i=0}^{n-1} x_i B^i,$$
where $x_i \in \mathbb{N}, \, x_i < B.$ In order words, square brackets mean zero-based tuple indexing.

@@

We will pretend natural numbers are in $\mathcal{F}$ by adopting the following identity:
$$n = \underbrace{\overline{1} + \overline{1} + \cdots + \overline{1}}_\text{$n$ times},$$
where $\overline{1} \in \mathcal{F}$ is the multiplicative identity in $\mathcal{F};$ the degenerate case of this identity is $0 = \overline{0},$
where $\overline{0} \in \mathcal{F}$ is the additive identity in $\mathcal{F}.$

@@

Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even prime,
the most widely used FFT algorithm, the Cooley---Tukey algorithm,
works by rewriting a composite $N = N_1 N_2$ into $N_1$ smaller transforms of size $N_2.$
Thus, it works only for the values of $N$ which are highly composite;
particularly when $N$ is a power of two.

@@

The main drawback of this algorithm is the precision issues.
Formally, it is not even a sound algorithm for multiplication: provided that the precision of
your floating-point values is bounded, you can not multiply arbitrarily large numbers with it: at
some point, you are going to get a round-off error large enough to produce a wrong digit in the
answer.

@@

A method intended to be faster, while not requiring a change in our representation of field
elements, is know as Barrett reduction. % TODO cite

@@

We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
$\approx$15\% compared to Montgomery reduction;
the code can be found in the \verb!bench-solinas! subdirectory of our repository.

@@

This leads us to the following algorithm, presented here as Python code:

\begin{verbatim}
def find_for_n(n, p_max, max_nyield):
    """
    Yields at most 'max_nyield' largest primes of form
      p = c * 3 * 2**n + 1,
    such that p <= p_max and 'c' is odd.
    """

    # Calculate the largest factor 'f' such that:
    #   (1) f is odd;
    #   (2) f is divisible by 3;
    #   (3) f * 2**n + 1 <= p_max.

    f = (p_max - 1) >> n
    if f % 2 == 0:
        f -= 1
    while f % 3 != 0:
        f -= 2

    # Search for primes of form
    #   (f - 6*i) * 2**n + 1,
    # where i >= 0.

    while f > 0 and max_nyield:
        p = (f << n) + 1
        if is_prime(p):
            yield p
            max_nyield -= 1
        f -= 6

def find_primes(n_min, w, ell):
    """
    Returns list of 'ell' largest primes of form
      p = c * 3 * 2**n + 1,
    where:
      (1) n_min <= n <= (w - 3);
      (2) p < 2**(w - 1);
      (3) 'c' is odd.
    """

    result = []
    p_max = (1 << (w - 1)) - 1
    for n in range(n_min, w - 2):
        result.extend(find_for_n(n, p_max, ell))
    assert len(result) >= ell
    return sorted(result)[-ell:]
\end{verbatim}

Note that it uses the external function \verb|is_prime|, which must perform primality testing.
We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
in \cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.

@@

The four-step algorithm is just a restatement, in terms of a $N_1 \times N_2$ matrix, of the
Cooley-Tukey algorithm that, in its general form, re-expresses a transform of length $N = N_1 N_2$
in terms of smaller transforms of lengths $N_1$ and $N_2.$
Thus, it can be used when it is desirable to perform a transform of length $N$ which is not a
power of two, so that the decimation-in-time (DIT) or decimation-in-frequency (DIF) variants of the
Cooley-Tukey algorithm can not be employed directly.

The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchial memory,
including external-memory transforms and modern systems with multiple levels of cache.

Remember our definition of the cyclic convolution in terms of DFT and IDFT:
$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes scalar (element-wise) product.

Note that, for computation of $\star,$ the specific order in which the direct transform
$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
$\mathfrak{f}^{-1}$ expects its inputs to be in, does not matter; we only need these two orders
to match.
Indeed, $\cdot$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
This means that, for the purpose of computing the cyclic convolution, in both four-step and six-step
algorithms, the final transposition step in DFT and the initial transposition step in IDFT, can be
simultaneously omitted. Note this does not mean you can omit just one of them but not another!

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage[ruled]{algorithm2e}

@@

\section{Notation}

The following notation is employed throughout this document:
$$\langle x_0, \ldots, x_{n-1} \rangle [i] = x_i.$$
In order words, square brackets mean zero-based tuple indexing.

A ring is a set $\mathcal{R}$ equipped with two binary operations, $+$ and $\cdot,$ such that
$\langle \mathcal{R}, + \rangle$ is an abelian group,
$\cdot$ is associative and there an identity element with respect to $\cdot$ in $\mathcal{R},$
and $\cdot$ is both left distributive and right distributive with respect to $+.$

@@

@@

Let us now fix a base $B \in \mathbb{N}, \, B > 1,$ and introduce the following notation:
$$\myllangle x_0, \ldots, x_{n-1} \myrrangle = \sum\limits_{i=0}^{n-1} x_i B^i,$$
where $x_i \in \mathbb{N}, \, x_i < B.$

@@

We will pretend natural numbers are in $\mathcal{F}$ by adopting the following identity:
$$n = \underbrace{\widehat{1} + \widehat{1} + \cdots + \widehat{1}}_\text{$n$ times},$$
where $\widehat{1} \in \mathcal{F}$ is the multiplicative identity in $\mathcal{F};$ the degenerate case of this identity is $0 = \widehat{0},$
where $\widehat{0} \in \mathcal{F}$ is the additive identity in $\mathcal{F}.$

@@

Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even prime,
the most widely used FFT algorithm, the Cooley---Tukey algorithm, works by re-writing a transform
of composite length $N = N_1 N_2$ into smaller transforms of lengths $N_1$ and $N_2.$
Thus, it works only for the values of $N$ which are highly composite;
particularly, if $N$ is a power of two, its variants named
decimation in time (DIT) and decimation in frequency (DIF), can be used.

@@

The main drawback of this algorithm is precision issues.
Formally, it is not even a sound algorithm for multiplication: provided that the precision of
your floating-point values is bounded, you can not multiply arbitrarily large numbers with it: at
some point, you are going to get a round-off error large enough to produce a wrong digit in the
answer.

@@

A method intended to be faster, while not requiring a change in our representation of field
elements, is known as Barrett reduction. % TODO cite

@@

We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
$\approx$ 15\% compared to Montgomery reduction;
the code can be found in the \verb!bench-solinas! subdirectory of our repository.

@@

This leads us to the following algorithm:

\begin{procedure}[H]
\caption{FindPrimesForN($n$, $p_{\max}, \ell$)}
\SetKw{KwAnd}{and}
\DontPrintSemicolon
\KwIn{$n, p_{\max}, \ell.$}
\KwOut{Set of $\ell$ largest, or less if the total number is less than $\ell,$ primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p \le p_{\max}$ and $c$ is odd.}
\Begin{
    $r \longleftarrow \varnothing$\;
    $\psi \longleftarrow (p_{\max} - 1) \mydiv 2^n $\;
    \If{$\psi \mymod 2 = 0$}{
        $\psi \longleftarrow \psi - 1$\;
    }
    \While{$\psi \mymod 3 \ne 0$}{
        $\psi \longleftarrow \psi - 2$\;
    }
    \While{$\psi > 0$ \KwAnd $|r| < \ell$}{
        $p \longleftarrow \psi \cdot 2^n + 1$\;
        \If{$p$ is prime}{
            $r \longleftarrow r \cup \{p\}$\;
        }
        $\psi \longleftarrow \psi - 6$\;
    }
    \Return{$r$}\;
}
\end{procedure}

\begin{procedure}[H]
\caption{FindPrimes($w$, $n_{\min}$, $\ell$)}
\SetKw{Error}{error}
\DontPrintSemicolon
\KwIn{$w, n_{\min}, \ell.$}
\KwOut{Set of $\ell$ largest primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p < 2^{w-1},$ $n \ge n_{\min},$ and $c$ is odd.}
\Begin{
    $r \longleftarrow \varnothing$\;
    $p_{\max} \longleftarrow 2^{w-1}-1$\;
    \For{$n \longleftarrow n_{\min}$ \KwTo $w-2$}{
        $r \longleftarrow r \cup \mathrm{FindPrimesForN}(n, p_{\max}, \ell)$\;
    }
    \eIf{$|r| < \ell$}{
        \Error{cannot find $\ell$ primes with required properties}\;
    }{
        \Return{$\ell$ largest values of $r$}\;
    }
}
\end{procedure}

Note that it uses an unspecified algorithm to perform primality testing.
We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
in \cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.

\section{Cooley-Tukey optimizations}

Before discussing optimizations,
we need to define the decimation in time (DIT) and decimation in frequency (DIF) variants of the
Cooley-Tukey FFT algorithm.

Both are specializations of Cooley-Tukey for a power-of-two length, and both include the step of
applying the bit-reversal permutation, which we introduce below, to the array being transformed,
either at the very beginning or in the very end.

Both algorithms take two parameters, an array $A$ of field elements and a field element
$\omega = \xi^{\pm 1},$
and return an array $A'$ of length $|A'| = |A|$
such that $$A'[k] = \sum\limits_{i=0}^{|A| - 1} A[i] \omega^{ik}.$$
If we express this transform as a function $\tau(A,\omega),$ then DFT and IDFT can be expressed in
terms of $\tau$ as follows:
\begin{eqnarray*}
\mathfrak{f}(x) &=& \tau(x, \xi); \\
\mathfrak{f}^{-1}(x) &=& N^{-1} \tau(x, \xi^{-1}),
\end{eqnarray*}
where $N$ is the length of the transform.

\subsection{Bit-reversal permutation}

Bit-reversal permutation is defined for sequences of length $2^n$ as follows: element with
zero-based index $k$ is exchanged with element with zero-based index $\mathbf{rev}_n(k),$ where
$\mathbf{rev}_n(k)$ is defined as the unique number in $[0; 2^n - 1]$ whose length-$n$ binary
representation equals to the reversed length-$n$ binary representation of $k.$

We now give the definition of the procedure that performs bit-reversal permutation.

\begin{procedure}[H]
\caption{BitRevPermute($A$)}
\DontPrintSemicolon
\KwIn{Array $A$, $|A| = 2^n.$}
\KwOut{The result of applying the bit-reversal permutation to $A.$}
\Begin{
    $n \longleftarrow \log_2 |A|$\;
    \For{$k \longleftarrow 0$ \KwTo $2^n - 1$}{
        $\langle b_0, \ldots, b_{n-1} \rangle \longleftarrow$ bits of $k;$ $b_i \in \{0,1\}$ and $\sum_{i=0}^{n-1} b_i 2^i = k$\;
        $k' \longleftarrow \sum_{i=0}^{n-1} b_{n-i-1} 2^i$\;
        \If{$k < k'$}{
            {exchange $A[k]$ with $A[k']$}\;
        }
    }
    \Return{$A$}\;
}
\end{procedure}

\subsection{Decimation in time}

We now define the decimation in time (DIT) algorithm.
Note that it performs bit-reversal permutation as the first step.
$\widehat{1}$ denotes the multiplicative identity of the underlying field.

\begin{procedure}[H]
\caption{DecimationInTime($A$, $\omega$)}
\SetKw{KwBy}{by}
\DontPrintSemicolon
\KwIn{
    Array $A$ of field elements, $|A| = 2^n$.
    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
}
\KwOut{
    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
}
\Begin{
    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
    \For{$s \longleftarrow 1$ \KwTo $\log_2 |A|$}{
        $m \longleftarrow 2^s$\;
        $\omega_m \longleftarrow \omega^{|A| / m}$\;
        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
            $\varphi \longleftarrow \widehat{1}$\;
            \For{$j = 0$ \KwTo $m/2-1$}{
                $u \longleftarrow A[k + j]$\;
                $v \longleftarrow \varphi \cdot A[k + j + m/2]$\;
                $A[k + j] \longleftarrow u + v$\;
                $A[k + j + m/2] \longleftarrow u - v$\;
                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
            }
        }
    }
    \Return{$A$}\;
}
\end{procedure}

The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as ``butterfly'', or,
more specifically, this one is the Cooley-Tukey butterfly; it maps
$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
\begin{eqnarray*}
x' &=& x + \varphi \cdot y; \\
y' &=& x - \varphi \cdot y.
\end{eqnarray*}

\subsection{Decimation in frequency}

We now define the decimation in frequency (DIF) algorithm.
Note that it performs bit-reversal permutation as the last step.
As in the previous subsection, $\widehat{1}$ denotes the multiplicative identity of the underlying field.

\begin{procedure}[H]
\caption{DecimationInFrequency($A$, $\omega$)}
\SetKw{KwBy}{by}
\SetKw{KwDownTo}{downto}
\DontPrintSemicolon
\KwIn{
    Array $A$ of field elements, $|A| = 2^n$.
    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
}
\KwOut{
    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
}
\Begin{
    \For{$s \longleftarrow \log_2 |A|$ \KwDownTo $1$}{
        $m \longleftarrow 2^s$\;
        $\omega_m \longleftarrow \omega^{|A| / m}$\;
        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
            $\varphi \longleftarrow \widehat{1}$\;
            \For{$j = 0$ \KwTo $m/2-1$}{
                $u \longleftarrow A[k + j]$\;
                $v \longleftarrow A[k + j + m/2]$\;
                $A[k + j] \longleftarrow u + v$\;
                $A[k + j + m/2] \longleftarrow \varphi \cdot (u - v)$\;
                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
            }
        }
    }
    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
    \Return{$A$}\;
}
\end{procedure}

The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as the Gentleman-Sande butterfly; it maps
$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
\begin{eqnarray*}
x' &=& x + y; \\
y' &=& \varphi \cdot (x - y).
\end{eqnarray*}

\subsection{Omitting the bit-reversal permutation}

Remember we expressed the cyclic convolution in terms of DFT and IDFT as follows:
$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes scalar (element-wise) product.

Note that, for the computation of $(\_ \star \_),$ the specific order in which the direct transform
$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
$\mathfrak{f}^{-1}$ expects its inputs to be in, do not matter; we only need these two orders
to match.
Indeed, $(\_ \cdot \_)$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$

This means that, for the purpose of computing the cyclic convolution, we can use DIF for the direct
transform and DIT for the inverse transform, and simultaneously omit the final permutation step in
DIF and the initial permutation step in DIT.
But note this does not mean just one of them can be omitted but not the other!

\subsection{Pre-calculating the table of root powers}

\subsection{Separate factor-1 butterfly}

@@

The four-step algorithm is just a restatement, in terms of $N_1 \times N_2$ matrix, of the
Cooley-Tukey algorithm that, in its general form, re-expresses a transform of length $N = N_1 N_2$
in terms of smaller transforms of lengths $N_1$ and $N_2.$
Thus, it can be used when it is desirable to perform a transform of a length that is not a
power of two, so that the decimation in time (DIT) or decimation in frequency (DIF) variants of the
Cooley-Tukey algorithm can not be employed directly.

The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchical memory,
including external-memory transforms and modern systems with multiple levels of cache.

\subsection{Omitting matrix transpositions}

\subsection{Algorithm for matrix transposition}

\subsection{Optimizing the four-step algorithm for $N=3 \cdot 2^k$}

\subsection{Optimizing the six-step algorithm for $N=2^k$}

\section{Low-level details}

\subsection{Modular addition and subtraction}

\subsection{Answer recovery}

\subsection{Power table}

\subsection{Shenanigans with Montgomery reduction}

Remember the algorithm for computing $\redc^\ast(t) = R^{-1} t$
in $\mathrm{F}_p,$
assumes $$0 \le t < p \mu,$$ where $\mu = 2^w$ if the machine word is $w$ bits long.
We can then compute $\redc(x,y),$ where $0 \le x,y < p$, as $\redc^\ast(xy).$

Remember also that we require $p < \mu/2.$
